---
title: "Effect of Transmission Type on MPG"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
Although, it initally looked as though manual cars had much higher MPG than automatic after regressing
out car weight that relationship filped. However, the 0.95 regression coefficient given to manual cars in this
model was not signficantly different from 1 at 5%, thus the results are inconclusive.


## Data Prep
The data was split for easier visualization and variables were converted to factors where appropriate. This resulted in a new data set called “cars”, which contains all variables and treats the discrete vairables as
factors.

```{r data prep, message=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
cars_num<-mtcars%>%select(mpg, am, disp, hp, drat,wt,qsec)
cars_fac1<-mtcars%>%select(mpg, am, cyl,vs)
cars_fac2<-mtcars%>%select(mpg, am,gear,carb)
cars_num$am<-factor(cars_num$am, labels=c("Auto","Manual"))
cars_fac1$am<-factor(cars_fac1$am, labels=c("Auto","Manual"))
cars_fac2$am<-factor(cars_fac1$am, labels=c("Auto","Manual"))
cars_fac1$cyl<-factor(cars_fac1$cyl)
cars_fac1$vs<-factor(cars_fac1$vs, labels=c("v","s"))
cars_fac2$gear<-factor(cars_fac2$gear)
cars_fac2$carb<-factor(cars_fac2$carb)
cars<-cbind(cars_fac1,cars_fac2[,3:4],cars_num[,3:7])
```

## Exploratory Analysis

### Numeric variables
Figure 1 in the appendix shows that all numeric varibles have reasonably strong correlations with MPG.
However, as seen in table 1 of the appendix, only 3 of these five have significantly different means when they
are compared across transmission type. Of those 3 weight (wt) and engine displacment (disp) are very strongly correlated to each other (0.888) it makes sense that the size of the engine would have this sort of relationship with weight so both variables do not need to be included.  Consequently, engine displacement has been dropped in favor of car weight because the car weight is slighly more correlated with mpg.


### Discrete Variables
Looking at figure 2 in the appendix it is clear that the automatic cars tend to have more cylinders than
manual (cyl) and that those cylinders tend to be arranged in a V instead of a straight line (vs). Cars with
more cylinders and cylinders in a V shape both tend to have lower miles per gallon. Figure 3 also shows that
automatic cars are almost exclusively 3 gear cars and that cars with 3 gears have the worst mpg performance.


## Model Selection
Consequently, only 5 varibles remain: weight of the car in 1,000 pounds (wt), rear axel ration (drat), number of cylinders in the engine (cyl), arrangement of those cylinders (vs), and the number of gears in the car (gear). Initially, results were derived with a Poisson transformation but upon completion it was noticed that the variance did not increase with lamda, but instead remained constant.  Thus a Quasipoisson model was used in its place.
The significance of each of these varibles on the influence transmission type has on mpg can be examined using an Anova analysis with nested models which can be seen in tables 2-5 of the appendix.  The first analysis in Table 2 shows that both weight and the number of gears significantly influence the relationship between mpg and transmission type.  Table 3 shows that cylinder count can also explain additional variance on top of weight, and Table 4 shows that the same is true of the engine shape but neither of these explain quite as much as the number of gears.  Table 5 shows the rear axel ratio does not significantly explain any additional variance.

```{r fit, message=FALSE, warning=FALSE}
fit<-glm(mpg~am+wt+gear,family="quasipoisson",data=cars)
```

## Residuals
 Figure 4 in the appendix shows that the reidudals are relatively constant throughout different predicted values which is alright because we are using the Quasipoisson transformation instead of a standard Poisson.

## Conclusion

The most appropriate model appears to use only transmission type, the weight of the car, and the number of gears as the predictor variables. This model yields a coefficient of 0.92 suggesting that manual cars are slightly less fuel efficient than automatic once weight and the number of gears are held constant.

```{r coef, message=FALSE, warning=FALSE}
exp(coef(fit))
```

However, the confidence interval of this coefficient contains 1 meaning the null hypothesis cannot be rejected
and nothing can be said about the affects of transmission type on mpg.

```{r confint, message=FALSE, warning=FALSE}
exp(confint(fit))
```


## Apendix

### Table 1
```{r table1, message=FALSE, warning=FALSE}
out <- data.frame("var1" = character()
, "var2"=character()
,"auto.mean"=integer()
,"manual.mean"=integer()
, "t.value"=integer()
, "df"=integer()
, "p.value"=double()
,row.names=NULL
)
for(i in 3:length(colnames(cars_num)))
{
test<-t.test(cars_num[,i]~am, data=cars_num)

out2 <- data.frame("var1" = colnames(cars_num)[i]
, "var2" = "am"
,"auto.mean"=test$estimate[1]
,"manual.mean"=test$estimate[2]
, "t.value" = sprintf("%.3f", test$statistic)
, "df"= test$parameter
, "p.value" = sprintf("%.3f", test$p.value)
,row.names=NULL
)
out<-rbind(out, out2)
}
out
```

### Table 2
```{r table2, message=FALSE, warning=FALSE}
fit1<-glm(mpg~am,family="quasipoisson",data=cars)
fit2<-glm(mpg~am+wt,family="quasipoisson",data=cars)
fit3<-glm(mpg~am+wt+gear,family="quasipoisson",data=cars)
fit4<-glm(mpg~am+wt+gear+cyl,family="quasipoisson",data=cars)
fit5<-glm(mpg~am+wt+gear+cyl+vs,family="quasipoisson",data=cars)
fit6<-glm(mpg~am+wt+gear+cyl+vs+drat,family="quasipoisson",data=cars)
anova(fit1,fit2,fit3,fit4,fit5,fit6, test="Chisq")
```

### Table 3
```{r talbe3, message=FALSE, warning=FALSE}
fit1<-glm(mpg~am,family="quasipoisson",data=cars)
fit2<-glm(mpg~am+wt,family="quasipoisson",data=cars)
fit3<-glm(mpg~am+wt+cyl,family="quasipoisson",data=cars)
fit4<-glm(mpg~am+wt+cyl+gear,family="quasipoisson",data=cars)
anova(fit1,fit2,fit3,fit4, test="Chisq")
```

### Table 4
```{r table4, message=FALSE, warning=FALSE}
fit1<-glm(mpg~am,family="quasipoisson",data=cars)
fit2<-glm(mpg~am+wt,family="quasipoisson",data=cars)
fit3<-glm(mpg~am+wt+vs,family="quasipoisson",data=cars)
fit4<-glm(mpg~am+wt+vs+gear,family="quasipoisson",data=cars)
anova(fit1,fit2,fit3,fit4, test="Chisq")
```

### Table 5
```{r table5, message=FALSE, warning=FALSE}
fit1<-glm(mpg~am,family="quasipoisson",data=cars)
fit2<-glm(mpg~am+wt,family="quasipoisson",data=cars)
fit3<-glm(mpg~am+wt+drat,family="quasipoisson",data=cars)
fit4<-glm(mpg~am+wt+drat+gear,family="quasipoisson",data=cars)
anova(fit1,fit2,fit3,fit4, test="Chisq")
```

### Figure 1
```{r figure1, message=FALSE, warning=FALSE}
g=ggpairs(cars_num, lower = list(continuous="smooth",discrete="facetbar"), diag=list(continuous="densityDiag"))
g
```

### Figrue 2
```{r figure2, message=FALSE, warning=FALSE}
g2=ggpairs(cars_fac1, lower = list(combo="blank",discrete="blank"), diag=list(continuous="densityDiag"))
g2
```

### Figrue 3
```{r figure3, message=FALSE, warning=FALSE}
g2=ggpairs(cars_fac2, lower = list(combo="blank",discrete="blank"), diag=list(continuous="densityDiag"))
g2
```

### Figrue 4
```{r figure4, message=FALSE, warning=FALSE}
plot(fit, which=1)
```

